---
title: "Netcdf: combining spatial and temporal data with metadata"
author: "Dasapta Erwin Irawan, R. Willem Vervoort & Gene Melzack"
date: "19 January 2018"
output: beamer_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(tidyverse)
```

## Storing and sharing spatio-temporal data

* Spatial and temporal data are challenging to store  
* It is 2 or 3 dimensional and can even be 4 dimensional (for example groundwater data)  
* We could use complex spreadsheets to share data, or build interconnected text files (by site, by time)  
    * This is not necessarily easy to share, you would need an additional read me file to describe the data (i.e. it is not self-describing)  
    * Is not necessarily system independent (proprietry software)  
* Solution: NetCDF formats

## NetCDF: history and principles

* NetCDF is the most widely used file format in climate and global studies
    * Almost all Global climate change model data is in NetCDF
    * Specifically good for multi-dimensional arrays
* Key strenghts are that NetCDF files are *self-describing* and *machine independent*  
* Libraries and protocols are maintained by [\textcolor{blue}{Unidata}](https://www.unidata.ucar.edu/software/netcdf/)  
* The latest protocol is NetCDF4, but older versions NetCDF3 are still around  
* The typical file extension used for NetCDF files is **.nc**  
* We will use the package **`ncdf4`** in R to read and create NetCDF4 files 

## NetCDF: Self-describing

* Self-describing means that within the file the metadata are included
* Here is an example from ET data from [\textcolor{blue}{NCI Thredds server ET data}](http://geonetworkrr9.nci.org.au/geonetwork/srv/eng/catalog.search?node=srv#/metadata/f749f8c8-b1a3-4125-a7ea-178fd4a31307)
* This is output from the [\textcolor{blue}{AWRA model}](http://www.bom.gov.au/water/landscape/)

```{r readnc, output.size=1}
library(ncdf4)

nc_example <- nc_open("../exercise/netcdf_data/AWRA05_data.nc")
print(nc_example)
```
