---
title: "Netcdf: combining spatial and temporal data with metadata"
author: "Dasapta Erwin Irawan, R. Willem Vervoort & Gene Melzack"
date: "22 January 2018"
output: beamer_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/rver4657/Google Drive/ITB_Usyd_Project2017/workshop_opendata")
library(knitr)
library(tidyverse)
library(pander)
```

## Storing and sharing spatio-temporal data

* Spatial and temporal data are challenging to store  
* It is 2 or 3 dimensional and can even be 4 dimensional (for example groundwater data)  
* We could use complex spreadsheets to share data, or build interconnected text files (by site, by time)  
    * This is not necessarily easy to share, you would need an additional read me file to describe the data (i.e. it is not self-describing)  
    * Is not necessarily system independent (proprietry software)  
* Solution: NetCDF formats

## NetCDF: history and principles

* NetCDF is the most widely used file format in climate and global studies
    * Almost all Global climate change model data is in NetCDF
    * Specifically good for multi-dimensional arrays
* Key strenghts are that NetCDF files are *self-describing* and *machine independent*  
* Libraries and protocols are maintained by [\textcolor{blue}{Unidata}](https://www.unidata.ucar.edu/software/netcdf/)  
* The latest protocol is NetCDF4, but older versions NetCDF3 are still around  
* The typical file extension used for NetCDF files is **.nc**  
* We will use the package **`ncdf4`** in R to read and create NetCDF4 files 

## NetCDF: Self-describing

* Self-describing means that within the file the metadata are included
* Here is an example from ET data from [\textcolor{blue}{NCI Thredds server ET data}](http://geonetworkrr9.nci.org.au/geonetwork/srv/eng/catalog.search?node=srv#/metadata/f749f8c8-b1a3-4125-a7ea-178fd4a31307)
* The next slide is output from the [\textcolor{blue}{AWRA model}](http://www.bom.gov.au/water/landscape/)

## netcdf metadata AWRA model

```{r readnc, echo=T}
library(ncdf4)

nc_example <- nc_open("C:/Users/rver4657/ownCloud/working/SSEAC/opendataworkshop2017/exercise/netcdf_data/AWRA05_data.nc")
print(nc_example)
```

## NetCDF: Self-describing

* The meta data and the description of the data are included in the file
* Each variable has to have information about units and a description

```{r ncInfo, echo=T}
# Longitude
lon <- ncvar_get(nc_example,"lon")
nlon <- dim(lon)
head(lon)
# units of time
tunits <- ncatt_get(nc_example,"time","units")
tunits$value

```

## NetCDF: Machine independent

* Because the algorithms to write and read netcdf are written in C and maintained by [\textcolor{blue}{unitdata}](https://www.unidata.ucar.edu/software/netcdf/) it can be used with any operating system.  
* It is open source, so you can recompile if needed  
* However, most scripting languages have tools to extract and manage netcdf files.  

## An example: creating a netcdf file using R

* We will be using the palæo-channel dataset that we have introduced on day 1  
* To review: this dataset consists of samples taken on a single date on 38 locations on a transect  
* To start of, we will transfer the particle size data to a netcdf (see below)

```{r PSA_nc, echo = T, message = F}
PSAdata <- read_csv("OriginalDataFolder/Willem/Soilparticles.csv")
pander(PSAdata[1:5,1:5])
```

## Defining the spatial and temporal dimensions

```{r defineDims, echo = TRUE}
library(ncdf4)

# define dimensions
londim <- ncdim_def("lon","degrees_east",
                    as.double(unique(PSAdata$Long)))
latdim <- ncdim_def("lat","degrees_north",
                    as.double(unique(PSAdata$Lat))) 
timedim <- ncdim_def("time","days sine 2001-01-21",
                     as.double(1))
depthdim <- ncdim_def("Depth","m", 
                      as.double(unique(PSAdata$Depth_top)))
```

* We actually don't need the time dimension, but I am putting it in as an example

## define variables

```{r defineVars}
fillvalue <- 1e32
dlname <- "Percent Clay"
clay_def <- ncvar_def("clay_perc","%",
                      list(londim,latdim,timedim,depthdim),fillvalue,
                      dlname, prec="single")
dlname <- "Percent Silt"
silt_def <- ncvar_def("silt_perc","%",
                      list(londim,latdim,timedim,depthdim),fillvalue,
                      dlname, prec="single")
dlname <- "Percent Fine Sand"
fsand_def <- ncvar_def("fsand_perc","%",
                       list(londim,latdim,timedim,depthdim),fillvalue,
                       dlname,prec="single")
dlname <- "Percent Coarse Sand"
csand_def <- ncvar_def("csand_perc","%",
                       list(londim,latdim,timedim,depthdim),fillvalue,
                       dlname,prec="single")
dlname <- "Percent Fine Gravel"
fgravel_def <- ncvar_def("fgravel_perc","%",
                         list(londim,latdim,timedim,depthdim),fillvalue,
                         dlname,prec="single")
dlname <- "Percent Coarse Gravel"
cgravel_def <- ncvar_def("cgravel_perc","%",
                         list(londim,latdim,timedim,depthdim),fillvalue,
                         dlname,prec="single")
```


## create netCDF file and put arrays

```{r makenc}
ncfname <- "PCSoildata_ncdf4.nc"
ncout <- nc_create(ncfname,list(clay_def,silt_def,fsand_def,csand_def,
                                fgravel_def, cgravel_def),force_v4=T)
```

* Need to first put data into arrays


